---
title: "Sentiment Analysis on Trump's Tweets"
author: "Rose Cheng"
date: "2026-03-08"
output: pdf_document
---

```{r}
# Load required libraries
library(dplyr)
library(stringr)
library(tidyr)
library(tidytext)
library(lubridate)
library(ggplot2)
library(broom)
# Import datasets
pre <- read.csv("/Users/rose/Desktop/Trump Tweets 2009-2024-selected/tweets_01-08-2021.csv", stringsAsFactors = FALSE)
post <- read.csv("/Users/rose/Desktop/Trump Tweets 2009-2024-selected/trump_post_2021.csv", stringsAsFactors = FALSE)
```

```{r}
library(textdata)
#define the periods (post and pre) & fix date times in post
pre2 <- pre %>%
  transmute(
  id = as.character(id),
  text = as.character(text),
  favorites = as.numeric(favorites),
  retweets = as.numeric(retweets),
  isRetweet = as.character(isRetweet),
  datetime = suppressWarnings(ymd_hms(date, tz = "UTC")),
  period = "Before 2021"
)
post2 <- post %>%
  transmute(
  id = as.character(id),
  text = as.character(text),
  favorites = as.numeric(favorites),
  retweets = as.numeric(retweets),
  isRetweet = as.character(isRetweet),
  datetime = suppressWarnings(ymd_hms(posix_date, tz = "UTC")),
  period = "After 2021"
)
#combine both datasets
df <- bind_rows(pre2, post2) %>%
  mutate(
  period = factor(period, levels = c("Before 2021", "After 2021")),
  isRetweet = case_when(
  isRetweet %in% c("t", "TRUE", "True", "1")~ TRUE,
  isRetweet %in% c("f", "FALSE", "False", "0")~ FALSE,
  TRUE~ NA
)
)
# Feature engineering (remove NA's, outliers, & get lengths)
df_feat <- df %>%
  mutate(
  char_count = str_length(text),
  caps_count = str_count(text, "\\b[A-Z]{2,}\\b"),
  exclaim_count = str_count(text, fixed("!")),
  emphasis = (caps_count > 0 & exclaim_count > 0)
) %>%
#filter outliers
  filter(char_count < 1200) %>%
  filter(!is.na(favorites), !is.na(retweets), !is.na(text)) %>%
  filter(!is.na(isRetweet)) #remove NAs
#load sentiment lexicons - bing & afinn
  bing <- get_sentiments("bing")
  bing <- bing %>%
  filter(!word %in% c("trump", "donald", "maga"))
  afinn <- get_sentiments("afinn")
  afinn <- afinn %>%
  filter(!word %in% c("trump", "donald", "maga"))
#clean and tokenize (bing) text
df_bing <- df_feat %>%
  mutate(text_clean = str_remove_all(text, "https?://\\S+")) %>%
  unnest_tokens(word, text_clean) %>%
  inner_join(bing, by = "word") %>%
  count(id, period, sentiment, name = "n") %>%
  mutate(score = if_else(sentiment== "positive", n,-n)) %>%
  summarise(sentiment_bing = sum(score), .by = c(id, period))
#clean and tokenize (afinn) text
df_afinn <- df_feat %>%
  mutate(text_clean = str_remove_all(text, "https?://\\S+")) %>%
  unnest_tokens(word, text_clean) %>%
  inner_join(afinn, by = "word") %>%
  summarise(sentiment_afinn = sum(value), .by = c(id, period))
#merge sentiments back into original data
df_model <- df_feat %>%
  left_join(df_bing, by = c("id", "period")) %>%
  left_join(df_afinn, by = c("id", "period")) %>%
  mutate(
  sentiment_bing = replace_na(sentiment_bing, 0),
  sentiment_afinn = replace_na(sentiment_afinn, 0),
  log_fav = log1p(favorites),
  log_rt = log1p(retweets)
)
#fix isRetweet error
df_model <- df_model %>%
  mutate(
  isRetweet = factor(isRetweet, levels = c(FALSE, TRUE)),
  emphasis = factor(emphasis, levels = c(FALSE, TRUE))
)
```

## Testing lexicon accuracy using a testing sample of 100 tweets:
```{r}
set.seed(100)
#draw samples from both datasets to evaluate baseline afinn lexicon accuracy:
sample_twitter <- df_model %>%
  filter(period == "Before 2021") %>%
  sample_n(100)

sample_truth_social <- df_model %>%
  filter(period == "After 2021") %>%
  sample_n(100)

#combine them:
sample_posts <- bind_rows(sample_twitter, sample_truth_social)
sample_posts <- sample_posts %>%
  mutate(human_sentiment = NA)

#export for human validation:
#write.csv(sample_posts, "sentiment_validation_sample.csv", row.names = FALSE)

sample_posts_evaluated <- read.csv("/Users/rose/Desktop/STATS 140XP Final project/code/sentiment_validation_sample.csv")
sample_posts_evaluated <- sample_posts_evaluated %>%
  mutate(
    afinn_label = case_when(
      sentiment_afinn > 0 ~ "positive",
      sentiment_afinn < 0 ~ "negative",
      TRUE ~ "neutral"
    )
  )
accuracy <- mean(sample_posts_evaluated$human_sentiment == sample_posts_evaluated$afinn_label)
accuracy

#accuracy by sentiment:
sample_posts_evaluated %>%
  mutate(correct = human_sentiment == afinn_label) %>%
  group_by(human_sentiment) %>%
  summarise(accuracy = mean(correct))
```

## Testing lexicon accuracy using the same testing sample with extended AFINN:
```{r}
##create custom dictionary:
custom_words <- tibble(
  word = c(
    "crooked","rigged","hoax","witch","radical","corrupt",
    "tremendous","incredible","fantastic","beautiful","greatest","successful"
  ),
  value = c(
    -3,-3,-3,-2,-2,-3,   # negative scores
    3,3,3,2,3,2          # positive scores
  )
)

## check which words are missing from AFINN:
missing_words <- setdiff(custom_words$word, afinn$word)
custom_words_filtered <- custom_words %>%
  filter(word %in% missing_words)

## extend AFINN:
afinn_extended <- bind_rows(afinn, custom_words_filtered)

## evaluate accuracy using extended AFINN:
validation_afinn_ext <- sample_posts_evaluated %>%
  mutate(text_clean = str_remove_all(text, "https?://\\S+")) %>%
  unnest_tokens(word, text_clean) %>%
  inner_join(afinn_extended, by = "word") %>%
  group_by(id) %>%
  summarise(score_ext = sum(value), .groups = "drop")

## merge scores back:
validation_compare <- sample_posts_evaluated %>%
  left_join(validation_afinn_ext, by = "id")

## fix missing scores and create labels:
sample_post_ext <- validation_compare %>%
  mutate(
    score_ext = replace_na(score_ext, 0),
    afinn_ext_label = case_when(
      score_ext > 0 ~ "positive",
      score_ext < 0 ~ "negative",
      TRUE ~ "neutral"
    )
  )

## compute accuracy for extended AFINN:
accuracy_ext <- mean(sample_post_ext$human_sentiment == sample_post_ext$afinn_ext_label)
accuracy_ext

#accuracy by sentiment:
sample_post_ext %>%
  mutate(correct = human_sentiment == afinn_ext_label) %>%
  group_by(human_sentiment) %>%
  summarise(accuracy = mean(correct))

table(sample_post_ext$human_sentiment,
      sample_post_ext$afinn_ext_label)
```


## Data cleaning: Compute sentiment scores using extended AFINN, removing retweets(fav = 0), and compute z-scores per platform
```{r}
## Compute sentiment scores for the entire dataset using the extended AFINN:
afinn_scores <- df_model %>%
  mutate(text_clean = str_remove_all(text, "https?://\\S+")) %>%
  unnest_tokens(word, text_clean) %>%
  inner_join(afinn_extended, by = "word") %>%
  group_by(id) %>%
  summarise(sentiment_afinn_ext = sum(value), .groups = "drop")

## Merge scores back to the full dataset
df_model_ext <- df_model %>%
  left_join(afinn_scores, by = "id")

## Replace missing scores:
df_model_ext <- df_model_ext %>%
  mutate(sentiment_afinn_ext = replace_na(sentiment_afinn_ext, 0))

## Convert score to sentiment label
df_model_ext <- df_model_ext %>%
  mutate(
    sentiment_label_ext = case_when(
      sentiment_afinn_ext > 0 ~ "positive",
      sentiment_afinn_ext < 0 ~ "negative",
      TRUE ~ "neutral"
    ) 
  ) %>%
  select(-c(sentiment_bing, sentiment_afinn))

## removing retweets (favorites = 0):
df_model_ext <- df_model_ext %>%
  filter(favorites != 0) 

## Compute z-scores per platform:
df_model_ext <- df_model_ext %>%
  group_by(period) %>%
  mutate(
    rel_fav = (favorites - mean(favorites, na.rm = TRUE)) / sd(favorites, na.rm = TRUE),
    rel_rt  = (retweets - mean(retweets, na.rm = TRUE)) / sd(retweets, na.rm = TRUE)
  ) %>%
  ungroup()
```

## EDA with the cleaned dataset:
```{r}
# distribution of sentiment_afinn_ext:
ggplot(df_model_ext, aes(x = sentiment_afinn_ext, fill = sentiment_label_ext)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of Sentiment Scores",
    x = "Sentiment Score",
    y = "Number of Posts",
    fill = "Sentiment"
  ) +
  theme_minimal()

# distribution of sentiment categories:
ggplot(df_model_ext, aes(x = sentiment_label_ext, fill = sentiment_label_ext)) +
  geom_bar() +
  labs(
    title = "Distribution of Sentiment Categories",
    x = "Sentiment",
    y = "Number of Posts"
  ) +
  theme_minimal()

## Distribution of relative engagement (Favorites)
ggplot(df_model_ext, aes(x = rel_fav)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Distribution of Relative Favorites",
    x = "Relative Favorites",
    y = "Count"
  ) +
  theme_minimal()

ggplot(df_model_ext, aes(x = rel_rt)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Distribution of Relative Retweets",
    x = "Relative Retweets",
    y = "Count"
  ) +
  theme_minimal()

# relative favorites by sentiment:
ggplot(df_model_ext, aes(x = sentiment_label_ext, y = log(rel_fav + 1), fill = sentiment_label_ext)) +
  geom_boxplot(outlier.alpha = 0.1) +
  labs(
    title = "Relative Favorites by Sentiment",
    x = "Sentiment",
    y = "Relative Favorites"
  ) +
  theme_minimal()

## relative retweet by sentiment
ggplot(df_model_ext, aes(x = sentiment_label_ext, y = log(rel_rt + 1), fill = sentiment_label_ext)) +
  geom_boxplot(outlier.alpha = 0.1) +
  labs(
    title = "Relative Retweets by Sentiment",
    x = "Sentiment",
    y = "Relative Retweets"
  ) +
  theme_minimal()

## sentiment–engagement(fav) relationship before vs after 2021
ggplot(df_model_ext, aes(sentiment_label_ext, log(rel_fav + 1), fill = period)) +
  geom_boxplot(outlier.alpha = 0.1) +
  labs(
    title = "Engagement by Sentiment Before vs After 2021",
    x = "Sentiment",
    y = "Relative Favorites"
  ) +
  theme_minimal()

## sentiment–engagement(retweets) relationship before vs after 2021
ggplot(df_model_ext, aes(sentiment_label_ext, log(rel_rt + 1), fill = period)) +
  geom_boxplot(outlier.alpha = 0.1) +
  labs(
    title = "Engagement by Sentiment Before vs After 2021",
    x = "Sentiment",
    y = "Relative Retweets"
  ) +
  theme_minimal()
```

## Linear regression model for research quesition 1:
```{r}
#model: log(favorites) = sentiment + controls
lm_fav1 <- lm(log(rel_fav + 1)~ sentiment_afinn_ext + period +  char_count + caps_count + exclaim_count + emphasis, data = df_model_ext)
tidy(lm_fav1, conf.int = TRUE)
summary(lm_fav1)
confint(lm_fav1)

#model: log(retweets) = sentiment + controls
lm_rt1 <- lm(log(rel_rt + 1)~ sentiment_afinn_ext + period + char_count + caps_count + exclaim_count + emphasis, data = df_model_ext)
tidy(lm_rt1)
summary(lm_rt1)
confint(lm_fav1)
```

## Linear regression model for research quesition 2:
```{r}
#model: log(favorites) = sentiment + period + sentiment x period + controls
lm_fav2 <- lm(log(rel_fav + 1) ~ sentiment_afinn_ext * period + char_count + caps_count + exclaim_count + emphasis, data = df_model_ext)
tidy(lm_fav2, conf.int = TRUE)
summary(lm_fav2)
confint(lm_fav2)

#model: log(retweets) = sentiment + period + sentiment x period + controls
lm_rt2 <- lm(log(rel_rt + 1) ~ sentiment_afinn_ext * period + char_count + caps_count + exclaim_count + emphasis, data = df_model_ext)
tidy(lm_rt2, conf.int = TRUE)
summary(lm_rt2)
confint(lm_rt2)
```

## interaction plot:
```{r}
## interaction between sentiment scores and fav:
library(patchwork)
df_model_ext$pred_rel_fav <- predict(lm_fav2)

sent_seq <- seq(min(df_model_ext$sentiment_afinn_ext),
                max(df_model_ext$sentiment_afinn_ext),
                length.out = 100)

pred_data <- expand.grid(
  sentiment_afinn_ext = sent_seq,
  period = c("Before 2021", "After 2021"),
  char_count = mean(df_model_ext$char_count),
  caps_count = mean(df_model_ext$caps_count),
  exclaim_count = mean(df_model_ext$exclaim_count),
  emphasis = factor(FALSE, levels = levels(df_model_ext$emphasis))
)

pred_data$pred <- predict(lm_fav2, newdata = pred_data)

p3 <- ggplot(pred_data,
       aes(x = sentiment_afinn_ext,
           y = pred,
           color = period)) +
  geom_line(size = 1.2) +
  labs(
    title = "Effect of Sentiment on Engagement Before and After 2021",
    x = "Sentiment Score",
    y = "Predicted Log(Relative Favorites + 1)"
  ) +
  theme_minimal()

## interaction between sentiment scores and retweets:
# sequence of sentiment values
sent_seq <- seq(min(df_model_ext$sentiment_afinn_ext),
                max(df_model_ext$sentiment_afinn_ext),
                length.out = 100)

# create prediction dataset
pred_data_rt <- expand.grid(
  sentiment_afinn_ext = sent_seq,
  period = c("Before 2021", "After 2021"),
  char_count = mean(df_model_ext$char_count),
  caps_count = mean(df_model_ext$caps_count),
  exclaim_count = mean(df_model_ext$exclaim_count),
  emphasis = factor(FALSE, levels = levels(df_model_ext$emphasis))
)

# predicted values from retweet model
pred_data_rt$pred <- predict(lm_rt2, newdata = pred_data_rt)

# plot
p4 <- ggplot(pred_data_rt,
       aes(x = sentiment_afinn_ext,
           y = pred,
           color = period)) +
  geom_line(size = 1.2) +
  labs(
    x = "Sentiment Score",
    y = "Predicted Log(Relative Retweets + 1)"
  ) +
  theme_minimal()

p3 + p4
```

## visualizing the correlation between sentiment score and engagment:
```{r}
ggplot(df_model_ext, aes(x = sentiment_afinn_ext, y = log(rel_fav + 1))) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "Sentiment vs Favorites",
    x = "Sentiment Score",
    y = "log(Favorites)"
  )

ggplot(df_model_ext, aes(x = sentiment_afinn_ext, y = log(rel_rt + 1))) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Sentiment vs Retweets",
    x = "Sentiment Score",
    y = "log(Retweets)"
  )

library(patchwork)

p1 <- ggplot(df_model_ext, aes(x = sentiment_afinn_ext, y = log(rel_fav + 1))) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "lm") +
  labs(title="Favorites")

p2 <- ggplot(df_model_ext, aes(x = sentiment_afinn_ext, y = log(rel_rt + 1))) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "lm") +
  labs(title="Retweets")

p1 + p2
```

## more data visualizations:
```{r}
#plot - sentiment vs favorites
ggplot(df_model_ext, aes(x = sentiment_afinn_ext, y = log(rel_fav + 1), color = period)) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
  title = "Sentiment vs Favorites (log1p), by Period",
  x = "AFINN sentiment score",
  y = "log1p(Favorites)"
)
#plot - sentiment distribution
ggplot(df_model_ext, aes(x = sentiment_afinn_ext, fill = period)) +
  geom_density(alpha = 0.4) +
  labs(
  title = "Distribution of Sentiment Scores by Period",
  x = "AFINN Sentiment Score",
  y = "Density"
)
#top words by period (descending order)
words_period <- df_feat %>%
  mutate(text_clean = str_remove_all(text, "https?://\\S+")) %>%
  unnest_tokens(word, text_clean) %>%
  inner_join(bing, by = "word") %>%
  count(period, sentiment, word, sort = TRUE)
  top_words <- words_period %>%
  group_by(period, sentiment) %>%
  slice_max(n, n = 10) %>%
  ungroup() %>%
  mutate(word = reorder_within(word, n, interaction(period, sentiment)))
#plot - top sentiments by period
ggplot(top_words, aes(x = word, y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ period + sentiment, scales = "free") +
  coord_flip() +
  scale_x_reordered() +
  labs(
  title = "Top Sentiment Words by Period",
  x = "Word",
  y = "Frequency"
)
```